Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-09-01 ms/main.tex

Cultural transmission may be either viewed as vertical, horizontal or oblique:  vertical transmission occurs between parents and offspring, horizontal transmission occurs between individuals from the same generation, and oblique transmission occurs  to offspring from the generation to which their parents belong (i.e. from non-parental adults). 
Evolution under either of these transmission models can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.
Both \citet{woodcock2006significance} and \citet{lewin2017microbes} demonstrated that non-vertical transmission can help explain the evolution of cooperative behavior (the former using simulations with cultural transmission, the latter using a model where cooperation is mediated by microbes that manipulate their host's behavior.) 
Specifically, some of the analyses by \citet{lewin2017microbes} can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.

Here we hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation because it generates assortment between cooperative individuals.
To test this hypothesis we construct a model in which behavioral changes can be mediated by cultural transmission that can occur during social interactions.
For example, an individual that interacts with a cooperative individual might learn that cooperation is a valuable behavior, and will therefore teach its offspring to be cooperative. 
Our cultural evolution models include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not because it facilitates the generation of assortment, and that treatment of cooperation as a cultural trait, rather than a genetic one, can lead to a better understanding of its evolutionary dynamics.

@@

% TODO uncomment when Result 3 in terms of c is fixed
%We define new cost boundaries, $\hat\gamma_1$ and $\hat\gamma_2$,
%\begin{equation} \label{eq:cost_boundaries_v}
%\begin{aligned}
%\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
%\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
%\end{aligned}
%\end{equation}
%and define a vertical transmission threshold,
%\begin{equation} \label{eq:v_threshold}
%\hat v = \frac{T_B - T_A}{1-T_A} \,.
%\end{equation}

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the 
polynomial coefficients $\beta_1$ and $\beta_3$ (\autoref{eq:polynomial_coefficients}):
%cost boundaries $\hat\gamma_1$ and $\hat\gamma_3$ (\autoref{eq:cost_boundaries_v}) and the vertical threshold $\hat v$ (\autoref{eq:v_threshold}) :

% in beta
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}: 
	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
	if \emph{(v)} $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
	if \emph{(vii)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}:
	if \emph{(viii)}  $T_A>T_B$ and $\beta_3<\beta_1$.
\end{enumerate}

% in gamma - TODO fix
%\begin{enumerate}
%\item \emph{Fixation of cooperation}: 
%	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \alpha \frac{T}{1-T}$; or
%	if \emph{(ii)} $T_A>T_B$ and $c < \hat\gamma_1$; or 
%	if \emph{(iii.a)} $T_A<T_B$ and $v>\hat v$ and $c < \hat\gamma_2$; or
%	if \emph{(iii.b)} $T_A<T_B$ and $v<\hat v$ and $\hat\gamma_2 < c$.
%\item \emph{Fixation of the defection}: 
%	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \alpha \frac{T}{1-T}$; or 
%  if \emph{(v)} $T_A>T_B$ and $\hat\gamma_1 < c < \hat\gamma_2$; or 
%	if \emph{(vi)} $T_A<T_B$ and $\hat\gamma_1 < c$.
%\item \emph{Fixation of either phenotype depending on initial frequency}:
%  if \emph{(viii)} $T_A>T_B$ and $v > \hat{v}$ and $\hat\gamma_2 < c$.
%\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
%  if \emph{(vii.a)} $T_A < T_B$ and $v < \hat{v}$ and $\hat\gamma_1 < c < \hat\gamma_2$; or
%  if \emph{(vii.b)} $T_A < T_B$ and $v>\hat{v}$ and $\hat\gamma_1 < c$ and $\hat\gamma_2 < c$.
%\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation and evaluated this hypothesis using a deterministic discrete-time evolutionary model with fitnesses in the form of payoffs from a prisoner's dilemma game.
Under oblique and horizontal cultural transmission, a horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under horizontal and vertical cultural transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the assortment between cooperation and transmission (Result~\autoref{result:vert_hori}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are more complicated, but all outcomes can still occur, depending on parameter values (Result~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_B>T_A$) and the cost of cooperation and assortment ratio are intermediate (yellow areas \autoref{fig:results2}).
% TODO figure for result 3
% TODO something about v-hat

>>>>

Cultural transmission may be either viewed as vertical, horizontal or oblique:  vertical transmission occurs between parents and offspring, horizontal transmission occurs between individuals from the same generation, and oblique transmission occurs  to offspring from the generation to which their parents belong (i.e. from non-parental adults). 
Evolution under either of these transmission models can be be more rapid than under pure vertical transmission~\citep{cavalli1981cultural,lycett2008questions,ram2018evolution}.
Both \citet{woodcock2006significance} and \citet{lewin2017microbes} demonstrated that non-vertical transmission can help explain the evolution of cooperative behavior (the former using simulations with cultural transmission, the latter using a model where cooperation is mediated by microbes that manipulate their host's behavior.) 
Some of the analyses by \citet{lewin2017microbes} can be applied to cultural transmission, because models of cultural transmission are mathematically similar to those for transmission of infectious diseases~\citep{cavalli1981cultural}.
Importantly, \citet{woodcock2006significance} showed the significance of non-vertical transmission for the evolution of cooperation. In their model individuals play a prisoner's dilemma game.
They do not link between interaction and transmission (i.e. no assortment), but their model demonstrates that it is possible to sustain altruistic behaviour via cultural transmission for a significant duration of time in the absence of any direct biological propensity favouring this behaviour.

We therefore hypothesize that non-vertical cultural transmission can enhance the evolution of cooperation.
To test this hypothesis we suggest a model in which behavioral changes are mediated by cultural transmission that can occur during social interactions. That is, if there exists assortative meeting in the choice of the social interaction partner, or assortative learning in choice of the cultural transmission partner.
For example, if an individual interacts with a cooperative individual, it might learn that cooperation is an advantageous behavior and will be cooperative in the future. 
Our cultural evolution models include both vertical and non-vertical transmission of cooperation, and we investigate these models using mathematical analysis and simulations.  
Our results demonstrate that cultural transmission can facilitate the evolution of cooperation even when genetic transmission can not because it facilitates the generation of assortment, and that treatment of cooperation as a cultural trait, rather than a genetic one, can lead to a better understanding of its evolutionary dynamics.

@@

We define new cost boundaries, $\hat\gamma_1$ and $\hat\gamma_2$,
\begin{equation} \label{eq:cost_boundaries_v}
\begin{aligned}
\hat\gamma_1 = \frac{b v \alpha T_A + (T_A - T_B)}{v(1-T_B)}, \quad
\hat\gamma_2 = \frac{b v \alpha T_B + (1+b) (T_A - T_B)}{v(1-T_B) + (1-v)(T_A-T_B)},
\end{aligned}
\end{equation}
and define a vertical transmission threshold,
\begin{equation} \label{eq:v_threshold}
\hat v = \frac{T_B - T_A}{1-T_A} \,.
\end{equation}

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
With vertical, horizontal, and oblique transmission, the cultural evolution of a cooperator phenotype will follow one of the following scenarios depending on the cost boundaries $\hat\gamma_1$ and $\hat\gamma_3$ (\autoref{eq:cost_boundaries_v}) and the vertical threshold $\hat v$ (\autoref{eq:v_threshold}) :

% in beta
%\begin{enumerate}
%\item \emph{Fixation of cooperation}: 
%	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
%	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
%	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
%\item \emph{Fixation of the defection}: 
%	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
%	if \emph{(v)} $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
%	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
%\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
%	if \emph{(vii)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
%\item \emph{Fixation of either phenotype depending on initial frequency}:
%	if \emph{(viii)}  $T_A>T_B$ and $\beta_3<\beta_1<0$.
%\end{enumerate}

% in gamma
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
	if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
	if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
	if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}: 
	if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
	if \emph{(v)} $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
	if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
  if \emph{(viia)} $T_A < T_B$ and $v<\hat{v}$ and $c<\hat{\gamma_1}$ and $c<\hat{\gamma_2}$; or
  if \emph{(viib)} $T_A < T_B$ and $v>\hat{v}$ and $c<\hat{\gamma_1}$ and $c>\hat{\gamma_2}$. %$0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}:
  if \emph{(viii)}  $T_A>T_B$ and $c > \hat{\gamma_2}$ and $c > \hat{\gamma_1}$ and $v>\hat{v}$.
\end{enumerate}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation and evaluated this hypothesis using a deterministic discrete-time evolutionary model with fitnesses in the form of payoffs from a prisoner's dilemma game.

Under oblique and horizontal cultural transmission, a horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under horizontal and vertical cultural transmission, cooperation or defection can either fix or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the assortment between cooperation and transmission (Result~\autoref{result:vert_hori}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are more complicated, but all outcomes can still occur, depending on parameter values (Result~\autoref{result:vert_obli_hori}).
Remarkably, stable coexistence between cooperation and defection can be maintained if horizontal transmission is biased for defection ($T_B>T_A$) and the cost of cooperation and assortment ratio are intermediate (yellow areas \autoref{fig:results2}).
% TODO figure for result 3
% TODO something about v-hat
